The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-19 dictionary.tex
Changes to Applications

Let $R$ complete discrete valuation ring $R$ with residue field $\Fq$.
Write $\m$ for the maximal ideal of $R$ and $K$ for the quotient field for $R$.
Set $R_n = R/\m^{n+1}$ and $S_n = \Spec{R_n}$ and $S = \Spec{R}$; 
note that $S_0 = \Spec{\Fq}$ is the special fibre of $S$.

@@

For any $n\in \NN_0$, we will write
\[
\Gr_n : \partop{\text{Schemes}}{\text{lft over $R$}} \to \partop{\text{Schemes}}{\text{lft over $\Fq$}}
\]
for the functor produced by composing pullback along $S_n \to S$ with the Greenberg functor. This functor respects \'etale morphisms, smooth morphisms and geometric components, \todo{open immersions? closed immersion?} and
\[
\Gr_n(X)(\Fq) \iso X(R_n),
\]
canonically, for any $X$ lft over $R$.

For any $n\leq m$,  the surjective ring homomorphism $R_{m} \to R_n$ determines a
natural transformation $\Gr_{m} \to \Gr_n$ between additive functors.
Crucially, $(\Gr_{m}\to \Gr_n)(X)$ is an affine morphism of $\Fq$-schemes, for every $R$-scheme $X$ and every $n\leq m$ \cite{bertrapelle-gonzales:Greenberg}*{Prop 4.3}.
This observation is key to the proof that, for any $X$ lft over $S$, the projective limit 
\[
\Gr(X) \ceq \varprojlim_{n\in \NN_0} \Gr_n(X)
\]
taken with respect to the surjective morphisms $(\Gr_{m}\to \Gr_n)(X)$,
exists in the category of group schemes over $\Fq$.
This leads to the definition of what we call the {\it Greenberg transform}:
\[
\Gr : \partop{\text{Schemes}}{\text{lft over $R$}} \to \partop{\text{Schemes}}{\text{lft over $\Fq$}}.
\]
By construction, the $\Fq$-scheme $\Gr(X)$ comes equipped with morphisms 
\[
P_n(X) : \Gr(X) \to \Gr_n(X),\qquad \forall n\in \NN_0.
\]
The Greenberg transform $\Gr$ respects \'etale morphisms, smooth morphisms  \todo{open immersions? closed immersion?} and geometric components, and
\[
\Gr(X)(\Fq) \iso X(R),
\]
canonically, for any $X$ lft over $R$.

@@

Now suppose $X_K$ is a commutative algebraic group over $K$ 
that admits a Neron model, $X$.
Then $\Gr(X)$ is commutative group scheme over $\Fq$
with component group scheme
$\pi_0(\Gr(X))= \pi_0(X) \times_S S_0$
and
\[
\Gr(X)(\Fq)\iso X_K(K)
\]
canonically.
For every $n\in \NN_0$,
$\Gr_n(X)$ is {\it smooth} commutative group scheme over $\Fq$
with component group scheme
$\pi_0(\Gr_n(X))= \pi_0(X) \times_S S_0$.

A {\it quasicharacter sheaf for $X_K$} is a triple
$\qcs{F}\ceq (\gqcs{F},\mu_\qcs{F},\phi_\qcs{F})$ where
\begin{enumerate}
\item[(0)] $\gqcs{F}$ is a constructible $\ell$-adic sheaf on $\Gr(X)_{\bFq} \ceq \Gr(X)\otimes \bFq$,
\item[(1)]  $\mu_\qcs{F}: \bm^* \gqcs{F} \to \gqcs{F}\boxtimes\gqcs{F}$ is an isomorphism $\ell$-adic sheaves, where $\bm : \Gr(X)_{\bFq} \times \Gr(X)_{\bFq} \to \Gr(X)_{\bFq}$ is multiplication for $\Gr(X)_{\bFq}$,
\item[(2)] $\phi_\qcs{F} : \Frob{\Gr(X)}^* \gqcs{F} \to \gqcs{F}$ is an
  isomorphism of constructible $\ell$-adic sheaves, where $\Frob{\Gr(X)}^* : \Gr(X)_{\bFq}  \to \Gr(X)_{\bFq}$ is Frobenius for $\Gr(X)_{\bFq}$,
\end{enumerate}
such that, for some $n\in \NN_0$,
\begin{enumerate}
\item[(3)]
$\qcs{F} = P_n(X)^*\qcs{L}$ for some quasicharacter sheaf $\qcs{L}$ on $\Gr_n(X)$, where pull-back by $P_n(X)$ is defined as in Proposition~\ref{prop:pullback}.
\end{enumerate}

\begin{theorem}
Let $X_K$ be a commutative algebraic group over $K$ with Neron model $X$ over $R$. 
\begin{enumerate}
\item[(0)] Quasicharacter sheaves for $X_K$ form a rigid monoidal category, $\QC(X_K)$, in a natural way. 
\item[(1)] There is a canonical short exact sequence of abelian groups 
\[
0 \to \Hh^2(\pi_0(X)_{\bFq},\EEx)^{\Frob{}} \to \QCiso{X_K} \to \Hom_\text{ad}(X_K(K),\EEx) \to 0
\] 
where $\Hom_\text{ad}(X_K(K),\EEx)$ is the group of characters of $X_K(K)$ that factor through $X(R)\to X(R_n)$, for some $n\in\NN_0$.
\item[(2)] For all quasicharacter sheaves $\qcs{F}$, $\qcs{F}'$ for $X_K$, and for every $\rho \in \Hom(\qcs{F},\qcs{F}')$, either $\rho$ is trivial or $\rho$ is an isomorphism.
\item[(3)] For all quasicharacter sheaves $\qcs{F}$ for $X_K$,
\[
\Aut(\qcs{L}) \iso \Hom((\pi_0(X)_{\Fq})_{\Weil{}},\EEx),
\]
canonically.
\end{enumerate}
\end{theorem}

>>>>

Let $R$ complete discrete valuation ring $R$ with residue field $\Fq$.
Write $\m$ for the maximal ideal of $R$ and $K$ for the quotient field for $R$.
Set $S = \Spec{R}$.
For every non-negative integer $n$, set $R_n = R/\m^{n+1}$ and $S_n = \Spec{R_n}$;  
note that $S_0 = \Spec{\Fq}$ is the special fibre of $S$.

@@

For any $n\in \NN_0$, we will write
\[
\Gr_n : \partop{\text{Schemes}}{\text{lft over $R$}} \to \partop{\text{Schemes}}{\text{lft over $\Fq$}}
\]
for the functor produced by composing pullback along $S_n \to S$ with the Greenberg functor. 
This functor respects open immersions, closed immersions, \'etale morphisms, smooth morphisms and geometric components, and
\[
\Gr_n(X)(\Fq) \iso X(R_n),
\]
canonically, for any $X$ lft over $R$.

For any $n\leq m$,  the surjective ring homomorphism $R_{m} \to R_n$ determines a
natural transformation $P_{n\leq m} : \Gr_{m} \to \Gr_n$ between additive functors.
Crucially, $P_{n\leq m}(X): \Gr_{m}(X)\to \Gr_n(X)$ is an affine morphism of $\Fq$-schemes, for every $R$-scheme $X$ and every $n\leq m$ \cite{bertrapelle-gonzales:Greenberg}*{Prop 4.3}.
This observation is key to the proof that, for any $X$ lft over $S$, the projective limit 
\[
\Gr(X) \ceq \varprojlim_{n\in \NN_0} \Gr_n(X)
\]
taken with respect to the surjective morphisms $P_{n\leq m}(X) : \Gr_{m}(X) \to \Gr_n(X)$,
exists in the category of group schemes over $\Fq$.
This leads to the definition of what we call the {\it Greenberg transform}:
\[
\Gr : \partop{\text{Schemes}}{\text{lft over $R$}} \to \partop{\text{Schemes}}{\text{lft over $\Fq$}}.
\]
By construction, the $\Fq$-scheme $\Gr(X)$ comes equipped with morphisms 
\[
P_n(X) : \Gr(X) \to \Gr_n(X),\qquad \forall n\in \NN_0.
\]
The Greenberg transform respects open immersions, closed immersions, \'etale morphisms, smooth morphisms and geometric components, and
\[
\Gr(X)(\Fq) \iso X(R),
\]
canonically, for any $X$ lft over $R$.

@@

Now suppose $X_K$ is a connected commutative algebraic group over $K$ with Neron model $X$ over $R$.
Then $\Gr(X)$ is commutative group scheme over $\Fq$
with component group scheme
\[
\pi_0(\Gr(X))= \pi_0(X) \times_S S_0
\]
and $\Fq$-rational points
\[
\Gr(X)(\Fq)\iso X_K(K)
\]
canonically.
For every $n\in \NN_0$,
$\Gr_n(X)$ is {\it smooth} commutative group scheme over $\Fq$
with component group scheme
$\pi_0(\Gr_n(X))= \pi_0(X) \times_S S_0$.

Consider the category $\mathcal{C}(\Gr(X))$ of triples
$\qcs{F}\ceq (\gqcs{F},\mu_\qcs{F},\phi_\qcs{F})$ where
\begin{enumerate}
\labitem{(C.0)}{C.0} $\gqcs{F}$ is a $\ell$-adic local system on $\Gr(X)_{\bFq} \ceq \Gr(X)\times_\Spec{\Fq} \Spec{\bFq}$,
\labitem{(C.1)}{C.1} $\mu_\qcs{F}: \bm^* \gqcs{F} \to \gqcs{F}\boxtimes\gqcs{F}$ is an isomorphism $\ell$-adic sheaves, where $\bm : \Gr(X)_{\bFq} \times \Gr(X)_{\bFq} \to \Gr(X)_{\bFq}$ is multiplication for $\Gr(X)_{\bFq}$,
\labitem{(C.2)}{C.2} $\phi_\qcs{F} : \Frob{\Gr(X)}^* \gqcs{F} \to \gqcs{F}$ is an
  isomorphism of constructible $\ell$-adic sheaves, where $\Frob{\Gr(X)}^* : \Gr(X)_{\bFq}  \to \Gr(X)_{\bFq}$ is Frobenius for $\Gr(X)_{\bFq}$,
\end{enumerate}
subject to the conditions as in \ref{QC.1} and \ref{QC.2},
with morphisms defined as in \ref{QC.3}.
This category is rigid monoidal, with tensor product defined as in Definition~\ref{def:QC}.

For every non-negative integer $n$, the morphism of $\Fq$-schemes $P_n(X) : \Gr(X) \to \Gr_n(N)$ induces a functor
\[
P_n(X)^* : \QC(\Gr_n(X)) \to \mathcal{C}(\Gr(X)),
\]
as in Proposition~\ref{prop:pullback}.
This functor is full and faithful, so we may view $\QC(\Gr_n(X))$ as a full subcategory of $\mathcal{C}(\Gr(X))$.

A {\it quasicharacter sheaf on $\Gr(X)$} is an object $\qcs{F}$ in $\mathcal{C}(\Gr(X))$ such that $\qcs{F} \iso P_n(X)^*\qcs{L}$ for some quasicharacter sheaf $\qcs{L}$ on $\Gr_n(X)$.
Quasicharacter sheaves on $\Gr(X)$ form a rigid monoidal category in the obvious way, denoted by $\QC(\Gr(X))$.
Note that we may view $\QC(\Gr(X))$ a `union' (colimit) of the subcategories $\QC(\Gr_n(X))$ in $\mathcal{C}(\Gr(X))$.

\begin{theorem}\label{thm:QCX}
Let $X_K$ be a commutative algebraic group over $K$ with Neron model $X$ over $R$. 
\begin{enumerate}
\item[(1)] There is a canonical short exact sequence of abelian groups 
\[
0 \to \Hh^2(\pi_0(X)_{\bFq},\EEx)^{\Frob{}} \to \QCiso{\Gr(X)} \to \Hom_\text{ad}(X_K(K),\EEx) \to 0
\] 
where $\Hom_\text{ad}(X_K(K),\EEx)$ is the group of homomorphisms $X_K(K)\to \EEx$ that factor through $X_K(X) = X(R)\to X(R_n)$, for some $n\in\NN_0$.
\item[(2)] For all quasicharacter sheaves $\qcs{F}$, $\qcs{F}'$ for $X_K$, and for every $\rho \in \Hom(\qcs{F},\qcs{F}')$, either $\rho$ is trivial or $\rho$ is an isomorphism.
\item[(3)] For all quasicharacter sheaves $\qcs{F}$ for $X_K$,
\[
\Aut(\qcs{F}) \iso \Hom((\pi_0(X)_{\Fq})_{\Weil{}},\EEx),
\]
canonically.
\end{enumerate}
\end{theorem}

\begin{proof}
To begin, fix $n\in \NN_0$. 
The following statements follow directly from Theorems~\ref{thm:snake} and \ref{thm:autornaught} together with our observations about the component group of $\Gr_n(X)$.
\begin{enumerate}
\item[(1)] There is a canonical short exact sequence of abelian groups 
\[
0 \to \Hh^2(\pi_0(X)_{\bFq},\EEx)^{\Frob{}} \to \QCiso{\Gr_n(X)} \to \Hom(X(R_n),\EEx) \to 0.
\] 
\item[(2)] For all quasicharacter sheaves $\qcs{L}$, $\qcs{L}'$ on $\Gr_n(X)$ and for every $\rho \in \Hom(\qcs{L},\qcs{L}')$, either $\rho$ is trivial or $\rho$ is an isomorphism.
\item[(3)] For all quasicharacter sheaves $\qcs{L}$ on $\Gr_n(X)$,
\[
\Aut(\qcs{L}) \iso \Hom((\pi_0(X)_{\Fq})_{\Weil{}},\EEx),
\]
canonically.
\end{enumerate}
So Theorem~\ref{thm:QCX} follows from 
\[
\QCiso{\Gr(X)} = \varinjlim_{n\in \NN_0} \QCiso{\Gr_n(X)}
\]
 and the definition of $\Hom_\text{ad}(X_K(K),\EEx)$.
\end{proof}

\subsection{Base change of quasicharacters}

\subsection{Transfer of quasicharacters}
